---
title: "Getting Started with multiRF"
author: "Wei Zhang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with multiRF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# Set eval_vignette = TRUE to execute all code chunks (recommended on a
# compute server).  The default is FALSE so that R CMD check / CRAN builds
# finish quickly without needing randomForestSRC to run.
eval_vignette <- identical(Sys.getenv("MULTIRF_EVAL_VIGNETTE"), "true") ||
  isTRUE(getOption("multiRF.eval_vignette"))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4.5,
  fig.align = "center",
  eval = TRUE,
  message = FALSE,
  warning = FALSE
)

library(multiRF)

data("tcga_brca_data")
```

> **Note:** Code chunks in this vignette are not evaluated during package
> build to keep install times short. To render with full output on a compute
> server, run:
>
> ```r
> # Option A: environment variable
> Sys.setenv(MULTIRF_EVAL_VIGNETTE = "true")
> rmarkdown::render("vignettes/multiRF.Rmd")
>
> # Option B: R option
> options(multiRF.eval_vignette = TRUE)
> rmarkdown::render("vignettes/multiRF.Rmd")
> ```

## Overview

`multiRF` provides a workflow for integrating matched multi-omics datasets with
random forests, MRF-style weighting, clustering, and downstream summaries.

The typical workflow is:

1. Prepare a named list of omics matrices with matched samples.
2. Run `mrf3()` to fit the integration model and produce cluster labels.
3. Inspect weights, selected variables, and cluster assignments.
4. Use plotting helpers for survival and embedding diagnostics.

The package exposes two main entry points:

- `mrf3()` for a compact end-to-end workflow with the most common arguments.
- `mrf3_fit()` for a stage-based workflow where IMD, variable selection, and
  robust clustering can be turned on explicitly.

## Method Maps

### Clustering choices

| Setting | Scope | Typical use |
|---------|-------|-------------|
| `similarity` | Global `main_clustering` | Default fused-similarity workflow; best starting point. |
| `proximity` | Global `main_clustering` | Use when RF proximity is the main object of interest. |
| `enhanced_proximity` | Global `main_clustering` | Aggressive proximity-based branch. |
| `PAM` | Shared or specific branch | Medoid-based clustering. |
| `Spectral` | Shared or specific branch | Similarity geometry-based clustering. |
| `Proximity` | Specific branch | Block-specific proximity alternative. |
| `Enhanced_Proximity` | Specific branch | Stronger proximity clustering for specific branches. |

### Variable-selection choices

| Method | Signal model | Typical use |
|--------|-------------|-------------|
| `filter` | Heuristic threshold | Good baseline for a practical first pass. |
| `test` | Significance-style | Hypothesis-testing flavor for weight selection. |
| `mixture` | Signal/noise mixture | When weights separate into background and enriched components. |

### Plotting choices

| Function | Input focus |
|----------|------------|
| `plot_tsne()` | Similarity or reconstruction |
| `plot_umap()` | Similarity or reconstruction |
| `plot_embed()` | Matrix-based quick embedding |
| `plot_network()` | Similarity graph |
| `plot_weights()` | IMD or cluster-IMD weights |
| `plot_circos()` | Adjacency-like structure |
| `plot_km()` | Cluster labels plus clinical |

## Example Data

This vignette uses the bundled TCGA BRCA data shipped with the package:

```{r example-data}
names(tcga_brca)
dim(tcga_brca$gene)
dim(tcga_brca$methy)
dim(tcga_brca$mirna)
```

Each block is a matrix-like data frame with samples in rows and features in
columns. Sample order should already be aligned across omics blocks before
fitting.

## Basic Clustering Workflow

For a standard clustering run, `mrf3()` is the simplest interface. It fits the
initial forests, learns block connections, reconstructs fused similarity
information, and returns cluster labels in a single call.

### Fitting the main workflow

The recommended entry point is `mrf3()`. We use `ntree = 50` here for speed
in vignette builds; in practice 300--500 trees are typical.

```{r fit-main, cache=TRUE}
fit <- mrf3(
  tcga_brca,
  k = 4,
  ntree = 50,
  filter_mode = "none",
  run_imd = TRUE,
  seed = 529
)
```

Once fitted, inspect the result:

```{r inspect-fit}
summary(fit)
table(get_clusters(fit))
names(get_weights(fit))
get_top_vars(fit, n = 10)
```

The returned object is an `mrf3_fit`. The most important fields for a basic
clustering analysis are:

- `fit$clusters`: main cluster assignments.
- `fit$connection`: learned directional block connections.
- `fit$model_top_v` and `fit$fused_top_v`: tuned sparsity levels.
- `fit$reconstruction`: reconstructed similarity and weight objects.
- `fit$shared` and `fit$specific`: shared/specific branch results.

### Choosing the main clustering backend

The global clustering backend is controlled by `main_clustering`. This setting
is propagated consistently through the shared, specific, and robust branches.

```{r clustering-backends, cache=TRUE}
fit_similarity <- mrf3(
  tcga_brca,
  k = 5,
  main_clustering = "similarity",
  ntree = 50,
  seed = 529
)

fit_proximity <- mrf3(
  tcga_brca,
  k = 5,
  main_clustering = "proximity",
  ntree = 50,
  seed = 529
)
```

### Comparing clusters to PAM50 subtypes

```{r cluster-vs-pam50}
common_ids <- intersect(
  names(get_clusters(fit)),
  tcga_brca_clinical$sampleID
)
pam50 <- tcga_brca_clinical$BRCA_Subtype_PAM50[
  match(common_ids, tcga_brca_clinical$sampleID)
]

cluster_metrics(get_clusters(fit)[common_ids], pam50)
```

You can also compare different global backends against each other:

```{r compare-backends}
fit_enhanced <- mrf3(
  tcga_brca,
  k = 5,
  main_clustering = "enhanced_proximity",
  ntree = 50,
  seed = 529
)

cluster_metric_matrix(
  list(
    similarity = get_clusters(fit_similarity)[common_ids],
    proximity = get_clusters(fit_proximity)[common_ids],
    enhanced_proximity = get_clusters(fit_enhanced)[common_ids],
    PAM50 = pam50
  ),
  metric = "ari"
)
```

```{r compare-backends-vs-pam50}
rbind(
  similarity = cluster_metrics(get_clusters(fit_similarity)[common_ids], pam50),
  proximity = cluster_metrics(get_clusters(fit_proximity)[common_ids], pam50),
  enhanced_proximity = cluster_metrics(get_clusters(fit_enhanced)[common_ids], pam50)
)
```

## Variable Selection Workflow

`mrf3_fit()` exposes optional stages such as IMD calculation and variable
selection more explicitly.

```{r vs-workflow, cache=TRUE}
fit_vs <- mrf3_fit(
  dat.list = tcga_brca,
  k = 5,
  ntree = 50,
  run_imd = TRUE,
  run_variable_selection = TRUE,
  variable_selection_args = list(
    method = "mixture",
    level = 0.05,
    normalized = TRUE
  ),
  seed = 529
)
```

Inspect the selected variables and summary:

```{r vs-inspect}
get_selected_vars(fit_vs)
get_vs_summary(fit_vs)
```

To compare different selection methods, reuse the fitted `fit_vs` object
which already contains IMD weights:

```{r vs-all-methods, cache=TRUE}
vs_filter <- mrf3_vs(
  mod = fit_vs,
  dat.list = tcga_brca,
  method = "filter",
  re_fit = FALSE,
  normalized = TRUE
)
vs_test <- mrf3_vs(
  mod = fit_vs,
  dat.list = tcga_brca,
  method = "test",
  re_fit = FALSE,
  level = 0.05
)
vs_mixture <- mrf3_vs(
  mod = fit_vs,
  dat.list = tcga_brca,
  method = "mixture",
  re_fit = FALSE,
  level = 0.05
)

lapply(
  list(
    filter = vs_filter,
    test = vs_test,
    mixture = vs_mixture
  ),
  get_selected_vars
)
```

## Robust and Sub-MRF Extensions

Robust clustering can be enabled within the staged workflow:

```{r robust-workflow, cache=TRUE}
fit_robust <- mrf3_fit(
  dat.list = tcga_brca,
  k = 5,
  ntree = 50,
  run_imd = TRUE,
  run_variable_selection = TRUE,
  run_robust_clustering = TRUE,
  variable_selection_args = list(method = "mixture"),
  seed = 529
)
```

```{r robust-inspect}
table(fit_robust$robust_clusters)
summary(fit_robust)
```

The sub-MRF mode runs multiple smaller forests and aggregates their behavior:

```{r submrf-workflow, cache=TRUE, eval=FALSE}
fit_sub <- mrf3_fit(
  dat.list = tcga_brca,
  k = 5,
  sub_mrf = TRUE,
  sub_mrf_args = list(
    n_sub = 10,
    frac_response = 0.2,
    frac_predictor = 0.2,
    ntree_per_sub = 5,
    compute_imd = TRUE
  ),
  run_imd = TRUE,
  seed = 529
)
```

```{r submrf-inspect, eval=FALSE}
summary(fit_sub)
table(get_clusters(fit_sub))
cluster_metrics(get_clusters(fit_sub)[common_ids], pam50)
```

## Plotting Helpers

The plotting helpers expose the main fitted structures without forcing you to
manually reconstruct the underlying matrices.

```{r plot-tsne, fig.width=7, fig.height=5}
plot_tsne(
  mod = fit,
  group = get_clusters(fit),
  source = "specific_shared",
  main = "t-SNE: shared/specific branch"
)
```

```{r plot-weights, fig.width=7, fig.height=5}
plot_weights(fit_vs, weight_source = "imd", top = 15)
```

```{r plot-km, fig.width=7, fig.height=6}
# Use sample names from the data to align with clinical
sample_ids <- rownames(tcga_brca[[1]])
clusters <- get_clusters(fit)
names(clusters) <- sample_ids

clinical_ids <- tcga_brca_clinical$sampleID
km_ids <- intersect(sample_ids, clinical_ids)
clinical_sub <- tcga_brca_clinical[match(km_ids, clinical_ids), , drop = FALSE]

plot_km(
  test_var = clusters[km_ids],
  time_var = "OS.time",
  event_var = "OS",
  pheno_mat = clinical_sub
)
```

## Practical Notes

- Start with `main_clustering = "similarity"` unless you specifically need
  proximity-driven behavior.
- Use lower `ntree` values only for prototyping or vignette builds.
- Turn on `run_variable_selection = TRUE` when interpretability is part of the
  goal, and compare multiple selection methods when signal is weak.
- Use `cluster_metrics()` and `cluster_metric_matrix()` to benchmark clustering
  choices against known labels or across methods.

## Session Info

```{r session-info}
sessionInfo()
```
