---
title: "Getting Started with multiRF"
author: "Wei Zhang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with multiRF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4.5,
  fig.align = "center",
  eval = FALSE,
  message = FALSE,
  warning = FALSE
)

library(multiRF)
data("tcga_brca", package = "multiRF", envir = knitr::knit_global())
data("tcga_brca_clinical", package = "multiRF", envir = knitr::knit_global())
```

## Overview

`multiRF` provides a workflow for integrating matched multi-omics datasets with
random forests, MRF-style weighting, clustering, and downstream summaries.

The typical workflow is:

1. Prepare a named list of omics matrices with matched samples.
2. Run `mrf3()` to fit the integration model and produce cluster labels.
3. Inspect weights, selected variables, and cluster assignments.
4. Use plotting helpers for survival and embedding diagnostics.

The package exposes two main entry points:

- `mrf3()` for a compact end-to-end workflow with the most common arguments.
- `mrf3_fit()` for a stage-based workflow where IMD, variable selection, and
  robust clustering can be turned on explicitly.

## Method Maps

### Clustering choices

| Setting | Scope | Typical use |
|---------|-------|-------------|
| `similarity` | Global `main_clustering` | Default fused-similarity workflow; best starting point. |
| `proximity` | Global `main_clustering` | Use when RF proximity is the main object of interest. |
| `enhanced_proximity` | Global `main_clustering` | Aggressive proximity-based branch. |
| `PAM` | Shared or specific branch | Medoid-based clustering. |
| `Spectral` | Shared or specific branch | Similarity geometry-based clustering. |
| `Proximity` | Specific branch | Block-specific proximity alternative. |
| `Enhanced_Proximity` | Specific branch | Stronger proximity clustering for specific branches. |

### Variable-selection choices

| Method | Signal model | Typical use |
|--------|-------------|-------------|
| `filter` | Heuristic threshold | Good baseline for a practical first pass. |
| `test` | Significance-style | Hypothesis-testing flavor for weight selection. |
| `mixture` | Signal/noise mixture | When weights separate into background and enriched components. |

### Plotting choices

| Function | Input focus | Typical question |
|----------|------------|-----------------|
| `plot_tsne()` | Similarity or reconstruction | Do clusters separate in a nonlinear embedding? |
| `plot_umap()` | Similarity or reconstruction | Is there robust manifold structure? |
| `plot_embed()` | Matrix-based quick embedding | Quick overview before deeper plotting? |
| `plot_heatmap()` | Similarity or fused block | Are samples or blocks visibly separated? |
| `plot_network()` | Similarity graph | Which sample relations dominate? |
| `plot_weights()` | IMD or cluster-IMD weights | Which features drive the result? |
| `plot_circos()` | Adjacency-like structure | How are modules connected globally? |
| `plot_km()` | Cluster labels plus clinical | Do clusters differ in survival? |

## Example Data

This vignette uses the bundled TCGA BRCA data shipped with the package:

```{r example-data}
names(tcga_brca)
dim(tcga_brca$gene)
dim(tcga_brca$methy)
dim(tcga_brca$mirna)
```

Each block is a matrix-like data frame with samples in rows and features in
columns. Sample order should already be aligned across omics blocks before
fitting.

## Basic Clustering Workflow

For a standard clustering run, `mrf3()` is the simplest interface. It fits the
initial forests, learns block connections, reconstructs fused similarity
information, and returns cluster labels in a single call.

### Fitting the main workflow

The recommended entry point is `mrf3()`. We use `ntree = 50` here for speed
in vignette builds; in practice 300--500 trees are typical.

```{r fit-main, cache=TRUE}
fit <- mrf3(
  tcga_brca,
  k = 4,
  ntree = 50,
  filter_mode = "none",
  run_imd = TRUE,
  seed = 529
)
```

Once fitted, inspect the result:

```{r inspect-fit}
summary(fit)
table(get_clusters(fit))
names(get_weights(fit))
get_top_vars(fit, n = 10)
```

The returned object is an `mrf3_fit`. The most important fields for a basic
clustering analysis are:

- `fit$clusters`: main cluster assignments.
- `fit$connection`: learned directional block connections.
- `fit$model_top_v` and `fit$fused_top_v`: tuned sparsity levels.
- `fit$reconstruction`: reconstructed similarity and weight objects.
- `fit$shared` and `fit$specific`: shared/specific branch results.

### Choosing the main clustering backend

The global clustering backend is controlled by `main_clustering`. This setting
is propagated consistently through the shared, specific, and robust branches.

```{r clustering-backends, cache=TRUE}
fit_similarity <- mrf3(
  tcga_brca,
  k = 4,
  main_clustering = "similarity",
  ntree = 50,
  seed = 529
)

fit_proximity <- mrf3(
  tcga_brca,
  k = 4,
  main_clustering = "proximity",
  ntree = 50,
  seed = 529
)
```

In practice:

- `"similarity"` is the default and the most natural choice when you want
  clustering based on fused similarity matrices.
- `"proximity"` uses random forest proximities more directly.
- `"enhanced_proximity"` is useful when you want a proximity-based branch with
  extra refinement.

## Comparing Cluster Assignments Against PAM50

The bundled clinical data includes PAM50 molecular subtypes, which provide a
natural external reference for evaluating cluster quality:

```{r cluster-vs-pam50}
common_ids <- intersect(
  names(get_clusters(fit)),
  tcga_brca_clinical$sampleID
)
pam50 <- tcga_brca_clinical$BRCA_Subtype_PAM50[
  match(common_ids, tcga_brca_clinical$sampleID)
]

# Compare main clusters vs PAM50
cluster_metrics(get_clusters(fit)[common_ids], pam50)
```

Compare multiple clustering solutions side by side:

```{r compare-backends}
cluster_metric_matrix(
  list(
    similarity = get_clusters(fit_similarity)[common_ids],
    proximity  = get_clusters(fit_proximity)[common_ids],
    PAM50      = pam50
  ),
  metric = "ari"
)
```

## Variable Selection Workflow

Variable selection is driven by IMD weights, so the usual pattern is to run IMD
and then enable `run_variable_selection = TRUE`.

```{r vs-workflow, cache=TRUE}
fit_vs <- mrf3_fit(
  dat.list = tcga_brca,
  k = 4,
  ntree = 50,
  run_imd = TRUE,
  run_variable_selection = TRUE,
  variable_selection_args = list(
    method = "mixture",
    level = 0.05,
    normalized = TRUE
  ),
  seed = 529
)
```

After this stage, the key outputs are:

- `fit_vs$selected_vars`: selected features per omics block.
- `fit_vs$selected_data`: filtered data matrices restricted to selected variables.
- `fit_vs$vs_summary`: block-level summary table for total vs selected features.

Typical inspection code:

```{r vs-inspect}
get_selected_vars(fit_vs)
get_vs_summary(fit_vs)
```

You can also call `mrf3_vs()` directly on an existing fit when you want to
experiment with different selection rules:

```{r vs-direct, cache=TRUE}
vs_filter <- mrf3_vs(
  mod = fit,
  dat.list = tcga_brca,
  method = "filter",
  re_fit = FALSE,
  normalized = TRUE
)

print(vs_filter)
```

### Running all methods side by side

```{r vs-all-methods, cache=TRUE}
vs_test <- mrf3_vs(mod = fit, dat.list = tcga_brca,
                   method = "test", re_fit = FALSE, level = 0.05)
vs_mixture <- mrf3_vs(mod = fit, dat.list = tcga_brca,
                      method = "mixture", re_fit = FALSE, level = 0.05)

lapply(
  list(filter = vs_filter, test = vs_test,
       mixture = vs_mixture),
  get_selected_vars
)
```

## Robust Clustering Workflow

Robust clustering first performs variable selection and then reclusters the
selected representation.

```{r robust-workflow, cache=TRUE}
fit_robust <- mrf3_fit(
  dat.list = tcga_brca,
  k = 4,
  ntree = 50,
  run_imd = TRUE,
  run_variable_selection = TRUE,
  run_robust_clustering = TRUE,
  variable_selection_args = list(method = "mixture"),
  seed = 529
)
```

```{r robust-inspect}
table(fit_robust$robust_clusters)
summary(fit_robust)
```

## Plotting Recipes

Most visualization helpers can work directly from an `mrf3_fit` object. The
`source` argument selects which matrix or branch to visualize.

### Weight plots

Weight plots are the quickest way to see which features dominate each block.

```{r plot-weights, fig.width=7, fig.height=5}
plot_weights(fit_vs, weight_source = "imd", top = 15)
```

### Heatmaps

```{r plot-heatmap, fig.width=7, fig.height=6}
plot_heatmap(
  dat = fit,
  group = get_clusters(fit),
  source = "specific_shared"
)
```

### Survival Analysis

`plot_km()` provides a convenience wrapper for Kaplan-Meier plots:

```{r survival, fig.width=6, fig.height=5}
clinical_ids <- tcga_brca_clinical$sampleID
common_ids <- intersect(names(get_clusters(fit)), clinical_ids)
clinical_sub <- tcga_brca_clinical[match(common_ids, clinical_ids), , drop = FALSE]

km <- plot_km(
  test_var = get_clusters(fit)[common_ids],
  time_var = "OS.time",
  event_var = "OS",
  pheno_mat = clinical_sub
)

km
```

## Recommended Workflow Patterns

For practical use, the following patterns are usually enough:

- Start with `mrf3()` when the goal is just clustering plus a quick read of
  the weights.
- Move to `mrf3_fit(run_imd = TRUE, run_variable_selection = TRUE)` when you
  want a reduced feature set.
- Add `run_robust_clustering = TRUE` when your final result should be based on
  selected variables rather than the initial full representation.
- Use the `source` argument in plotting functions to decide which branch or
  reconstruction object to visualize.

## Notes

- Keep sample order matched across omics blocks before fitting.
- Use `filter_omics()` for explicit pre-filtering before model fitting.
- For pairwise clustering comparisons, see `cluster_metrics()` and
  `cluster_metric_matrix()`.

## Session Info

```{r session-info}
sessionInfo()
```
